---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(ggplot2)
```
1. Read in `districts.csv` into a tibble named `districts`. How many rows and columns does it contain?
146 rows, 27 columns
```{r}
districts <- read.csv('../data/districts.csv')
glimpse(districts)
```

2. Notice that the first row corresponds to the whole State of Tennessee. Remove this row and save the result back to `districts`.
```{r}
districts <- slice(districts,(2:n()))
```

3. How many districts have a proficiency rate of at least 80% for both alg_1 and eng_1?
13
```{r}
filter(districts, alg_1 >= 80 & eng_1 >= 80) %>% count()
```
or
```{r}
count(districts, alg_1 >= 80 & eng_1 >= 80)
```


4. How many districts have a proficiency rate less than 50% for either alg_1 or eng_1?
8
```{r}
filter(districts, alg_1 < 50 | eng_1 < 50) %>% count()
```

5. Which district has the lowest graduation rate?
Tenn School for Blind
```{r}
districts %>% 
  summarize(system_name[which.min(grad)])
```

6. Which district has the highest dropout rate?
Achievement School District
```{r}
districts %>% summarize(system_name[which.max(dropout)])
```

7. Within the Mid Cumberland region, which district has the highest ACT composite?
Williamson County
```{r}
filter(districts, region == "Mid Cumberland") %>% 
    summarize(system_name[which.max(act_composite)])
```

8. Create a histogram showing the distribution of graduation rates. What can you say about this distribution?
The distribution is very right skewed. Almost 80 districts have a 90% or higher graduation rate. Roughly 
50 districts have a 80 percent or higher graduation rate. Approximately 10 districts have less than an 
80 percent graduation rate. Only one district has a less than 20 percent graduation rate.
```{r}
ggplot(districts, aes(x = grad)) +
  geom_histogram()
```


9. Create a scatter plot to compare alg_1 proficiency rates to alg_2 rates. What do you notice? Facet this plot by region. Does anything stand out when you facet the plots?
There appears to be a positive correlation between the scores for alg_1 and alg_2.
Some schools have a very narrow range of variance relative to others.


```{r}
ggplot(districts, aes(x = alg_1, y = alg_2)) +
  geom_point()
```


```{r}
ggplot(data = districts, aes(alg_1, alg_2)) +
  geom_point() +
  # labs( y = "Rating", x = "Percent") + 
  facet_grid( ~ region) 
  # facet_grid(rows = vars(region)) 
```

10. Create a bar chart showing the total enrollment by region. Which region has the highest total enrollment? Which has the smallest?
Highest enrollment: Mid Cumberland
Smallest enrollment: Northwest
```{r}
ggplot(districts, aes(region)) +
  geom_bar(aes(weight = enrollment)) + 
  theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1))
```

11. When creating this bar chart you may have noticed that some districts have missing enrollment values. For how many districts is this the case?
4
```{r}
filter(districts, is.na(districts$enrollment)) %>% count()
```


12. What is the mean graduation rate across all districts? What might be wrong with using just the regular mean to assess average graduation rates?
90%
The districts do not have the same number of students so some schools get a higher weight for the 
graduation rate.
```{r}
mean(districts$grad, na.rm = TRUE)
```

13. Redo the previous question but use a weighted average (`weighted.mean`) graduation across all districts, weighing by enrollment. How much does this change your answer? Can you explain using the data the reason for the big change from using the mean?
87.33%
```{r}
enroll_grad <- districts %>% select(enrollment, grad) %>% na.omit()
weighted.mean(enroll_grad$grad, enroll_grad$enrollment)
```


14. Create a boxplot showing enrollment rates per region. Does anything stand out?
Mid Cumberland has a much larger enrollment than any other region, 
Northwest a much lower enrollment that the other regions.

```{r}
ggplot(districts, aes(region, enrollment)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1))
```


15. Create a boxplot showing graduation rates per region. Does anything stand out?
Southwest/Memphis has the greatest range of graduation rates. Mid Cumberland has the lowest
overall graduation for one of its districts.

```{r}
ggplot(districts, aes(region, grad)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1))
```


16. Find the weighted average of graduation rates by region using enrollment as weights. Compare the results you get for the weighted average to what you see from the boxplots. Can you explain any discrepancy?
weighted average to what you see from the boxplots. Can you explain any discrepancy?
Southwest/Memphis has a much lower weighted graduation average than what shows in the box plot. This is probably
due to there being many more enrolled in the other districts that have a higher graduation rate.

```{r}
enroll_region_grad <- districts %>% select(enrollment, region, grad) %>% na.omit()
enroll_region_grad %>% 
  group_by(region) %>% 
  summarise(weighted_grad = weighted.mean(grad, enrollment))
```


17. For many districts, values for `alg_2` are lower than for `alg_1`. Create a histogram showing the distribution of differences (`alg_1` - `alg_2`). Which school had the largest drop from `alg_1` to `alg_2`? For what percentage of schools is it true that `alg_2` is larger than `alg_1`? Is there a similar dropoff for `eng_2` and `eng_3`?
Pickett County had the largest drop from alg_1 to alg_2.
For 15.4% of the schools alg_2 is larger than alg_1.
The shape of the curve for the histogram for the eng_2/eng_3 difference is much narrower and centered around the 20 percent
value.
The percent of districts with eng_3 greater than eng_2 is 0.8%.

```{r}
q17 <- function(col1, col2) {
  data <- districts %>% select(all_of("system_name"), all_of(col1), all_of(col2)) %>% 
    na.omit() %>% data.frame()
  data$col_diff <- (data[[col1]] - data[[col2]])
  val1 <- filter(data, data$col_diff == max(data$col_diff))$system_name
  diff_hist <- ggplot(data, aes(col_diff)) +
    geom_histogram()
  val2 <- filter(districts, districts[[col2]] - districts[[col1]] > 0) %>% count() / data %>% count() * 100
  return(list(val1, val2, diff_hist))
}
```
```{r}
q17("alg_1", "alg_2")
q17("eng_2", "eng_3")
```


18. You may have noticed that a lot of rows are missing values. Which district has the largest number of missing values? What do you notice about schools that have a lot of missing values?
West Tenn School for Deaf
Schools with a large number of missing values are for the most part all missing values for eng_2, eng_3, alg_2, bio, and chem.

```{r}
districts$na_count <- apply(is.na(districts), 1, sum)
filter(districts, districts$na_count == max(districts$na_count))
filter(districts, districts$na_count >= 9)
ggplot(districts, aes(na_count)) +  geom_bar()
```


19. Find the correlation between graduation rate and all other variables. Create a horizontal bar chart showing these correlations. Make sure that your plot is ordered by correlation values. What do you notice from these correlations?
The largest correlation rate is between dropout and graduation. Other variable have a smooth transition from a negative
correlation to a positive correlation but all correlation falling within the +/- .5 value.

```{r}
d_no_na <- districts %>% na.omit()
d_matrix <- data.matrix(d_no_na[3:27])
cor_df <- data.frame(cor(d_matrix[,23], d_matrix[, -c(23)]))
cor_df <- data.frame(name = colnames(cor_df), value = as.numeric(as.vector(cor_df[1,])))
ggplot(cor_df, aes(name)) +  geom_bar()
cor_df
```
```{r}
ggplot(cor_df, aes(x = reorder(name, -value), y = value)) +   
  geom_bar(stat = "identity") +
  coord_flip()
```


20. Create a scatterplot for `grad` vs. `suspended`. Does what you see make sense given your answer from the previous part?
This does make sense given the correlation values seen previously. There is a correlation abeit a weak one.

```{r}
ggplot(data = districts, aes(grad, suspended)) +
  geom_point() 
```


21. Create a linear regression model using `lm` with target variable `grad` and predictor variable `suspended`. What R^2 value does this model have? What is the interpretation of this number?
0.0795
There is very little predictive ability from the suspended rate on the graduation rate.

```{r}
simple_regression_model <- lm(grad ~ suspended, data = districts)

summary(simple_regression_model)

# ggplot(districts, aes(x= suspended, y = grad)) + geom_point()
```


22. Add the regression line to your scatterplot using `geom_smooth` with `method='lm'`. How do you feel about the regression line after seeing it plotted on the scatterplot?
The regression line seems about right as there is very little correlation between 
suspended and graduation rates.

```{r}
ggplot(data = districts, aes(grad, suspended)) +
  geom_point() +
  geom_smooth(method='lm')
```


**Continued Exploration and Practice**

23. Read in the school-level testing data for 2014, available [here](https://www.tn.gov/content/dam/tn/education/data/data_2014_school_base.xlsx). You might find the readxl library useful for this task. If you use this library, be sure to look at the `na` argument for the `read_excel` function.

```{r}
base <- readxl::read_excel('../data/data_2014_school_base.xlsx', na = "*")

```

24. How many schools have at least 20 percent of students below bsc for Algebra I? Which districts do these schools belong to? 
71 schools have at least 20 percent of students below bsc for Algebra I
 [1] "Bedford County"              "Bledsoe County"             
 [3] "Maryville"                   "Bradley County"             
 [5] "Campbell County"             "Hollow Rock-Bruceton"       
 [7] "Cheatham County"             "Claiborne County"           
 [9] "Cumberland County"           "Davidson County"            
[11] "Fayette County"              "Giles County"               
[13] "Grainger County"             "Hamilton County"            
[15] "Haywood County"              "Hickman County"             
[17] "Jackson County"              "Knox County"                
[19] "Lake County"                 "Lewis County"               
[21] "Lincoln County"              "Fayetteville"               
[23] "Jackson-Madison County"      "Morgan County"              
[25] "Polk County"                 "Sevier County"              
[27] "Shelby County (New)"         "Weakley County"             
[29] "Tenn School For Blind"       "Achievement School District"

```{r}
base %>% 
  filter(pct_below_bsc != "NA" & pct_below_bsc != "**") %>%
  mutate(pct_below_bsc = as.numeric(pct_below_bsc)) %>%
  filter(grade == 'All Grades' & subgroup == 'All Students' &
                  subject == "Algebra I" & pct_below_bsc >= 20) %>% count()
```
```{r}
base %>% 
  filter(pct_below_bsc != "NA" & pct_below_bsc != "**") %>%
  mutate(pct_below_bsc = as.numeric(pct_below_bsc)) %>%
  filter(grade == 'All Grades' & subgroup == 'All Students' &
                  subject == "Algebra I" & pct_below_bsc >= 20) %>% 
  distinct(system_name) %>% pull()
```


25. How many schools have at least 20 percent of students below bsc for _both_ Algebra I and English I?
4

```{r}
alg_below <- base %>% 
  filter(pct_below_bsc != "NA" & pct_below_bsc != "**") %>%
  mutate(pct_below_bsc = as.numeric(pct_below_bsc)) %>%
  filter(grade == 'All Grades' & subgroup == 'All Students' &
                  subject == "Algebra I" & pct_below_bsc >= 20) 

eng_below <- base %>% 
  filter(pct_below_bsc != "NA" & pct_below_bsc != "**") %>%
  mutate(pct_below_bsc = as.numeric(pct_below_bsc)) %>%
  filter(grade == 'All Grades' & subgroup == 'All Students' &
                  subject == "English I" & pct_below_bsc >= 20) 
inner_join(alg_below, eng_below, by = c("system_name", "school_name")) %>%
  select(system_name, school_name)
```


26. Which grade has the highest pct_adv for Algebra I? Plot the average pct_adv per grade level as a bar chart. Make sure that the bars are ordered by grade level.
8th grade has the highest pct_adv for Algebra I.

```{r}
alg_adv <- base %>% filter(grade %in% (8:12) & subject == "Algebra I" & subgroup == 'All Students') %>% 
  group_by(grade) %>% summarise(value = mean(as.numeric(pct_adv), na.rm = TRUE))
alg_adv <- transform(alg_adv, grade = as.numeric(grade))
```

```{r}
alg_adv %>%
  ggplot(aes(x = grade, value, y = value)) +   
    geom_bar(stat = "identity")
```


27. Find the correlation between pct_adv for Algebra I and pct_adv for Algebra II by school. Create a scatterplot showing Algebra II scores vs. Algebra I scores by school.
0.603

```{r}
alg1_pct <- base %>% 
  filter(pct_adv != "NA" & pct_adv != "**") %>%
  mutate(pct_adv = as.numeric(pct_adv)) %>%
  filter(subject == "Algebra I", grade == "All Grades", subgroup == "All Students") 
alg2_pct <- base %>% 
  filter(pct_adv != "NA" & pct_adv != "**") %>%
  mutate(pct_adv = as.numeric(pct_adv)) %>%
  filter(subject == "Algebra II", grade == "All Grades", subgroup == "All Students")
alg_inner <- inner_join(alg1_pct, alg2_pct, by = c("system_name", "school_name"))
cor(alg_inner$pct_adv.x, alg_inner$pct_adv.y)
```

```{r}
alg_inner %>%
  ggplot(aes(x = pct_adv.x, y = pct_adv.y)) +
  geom_point()
```


28. Find all schools in Rutherford County that have "High School" in their name. For these schools, create a chart (your choice) showing the differences in pct_below_bsc, pct_bsc, pct_prof, and pct_adv for Algebra I when looking across all subgroups and grades.

```{r}
rutherford_hs <- base %>%  
  filter(system_name == "Rutherford County" & grepl("High School", school_name) &
           grade %in% (9:12) ) %>%
    mutate(pct_below_bsc = as.numeric(pct_below_bsc), pct_bsc = as.numeric(pct_bsc), 
           pct_prof = as.numeric(pct_prof), pct_adv = as.numeric(pct_adv),
           grade = as.numeric(grade)) 

```

```{r}
rutherford_hs %>%
  ggplot( ) +
  geom_point(aes(x = pct_below_bsc, y = pct_bsc), color = 'green') +
  geom_point(aes(x = pct_prof, y = pct_adv), color = 'blue') +

  theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1)) +
  facet_grid(vars(grade), vars(subgroup))

```


29. I claim that smaller schools do a better job preparing students for Algebra I standardized tests. Find the average number of valid tests (a proxy for the school size) for schools where the pct_prof_adv for Algebra I is greater than 95. Compare this to the average number of valid tests for all schools. In light of this result, how does my claim look?
From the comparison of average number of tests where the pct_prof_adv is above 95, which is 48.4, and the average number of tests overall, 130.0, a smaller the better tests scores do indeed seem to be in smaller schools.

```{r}
base %>%
  filter(subject == "Algebra I", pct_prof_adv > 95, grade == "All Grades", subgroup == "All Students") %>%
  pull(valid_tests) %>% mean()
base %>%
  filter(subject == "Algebra I", grade == "All Grades", subgroup == "All Students") %>%
  pull(valid_tests) %>% mean()  
```


30. I also claim that smaller schools do a worse job preparing students for Algebra I standardized tests. Find the average number of valid tests (a proxy for the school size) for schools where the pct_prof_adv for Algebra I is less than 25. Compare this to the average number of valid tests for all schools. In light of this result, how does my claim look now?
From the average number of tests for pct_prof_adv below 25, which is 62.1, compared to the average test number for all schools, which is 130.0, smaller schools do indeed seem to do worse at preparing students for the Algebra I standardized tests.

```{r}
base %>%
  filter(subject == "Algebra I", pct_prof_adv < 25, grade == "All Grades", subgroup == "All Students") %>%
  pull(valid_tests) %>% mean()
base %>%
  filter(subject == "Algebra I", grade == "All Grades", subgroup == "All Students") %>%
  pull(valid_tests) %>% mean()
```


31. Create a scatterplot showing pct_prov_adv vs. valid_tests. Can you use this to explain the result for numbers 26 and 27?
The lower and higher side of the pct_prof_adv scale are lower on the valid_schools scale. 

```{r}
base %>%
  filter(subject == "Algebra I", grade == "All Grades", subgroup == "All Students") %>%
  ggplot() +
  geom_point( aes(x = pct_prof_adv, y = valid_tests))
```


If you finish all of the above questions, continue to explore the two datasets and see what else interesting you can find.

Also, check out the plotly library for R. The `ggplotly` function makes it very easy to convert ggplot plots into interactive plotly plots.

